Improved Hierarchical Clustering Through Nesting

### 2015-05-30 icdm/submission.tex
Results and updated submission

Recall the hierarchical $k$-median problem: given a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$, we will represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$.
 A solution to the hierarchical $k$-median problem is a chain of nested solutions $(S_1, a_1) \subseteq (S_2, a_2) \subseteq \ldots \subseteq (S_l, a_l)$. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the hierarchy.  
Consider any $k$-median algorithm $\mathcal{A}$. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce an augmentation step that will nest these clusterings into the hierarchical framework and only increase costs by a bounded factor.

First, set $i=1$, $S_0 = \emptyset$, $\beta = 3+\sqrt{3}$, and $\beta_0 = 1$ (for the randomized algorithm we switch $\beta = 6.355$ and set $\beta_0 = \beta^X$, where $X$ is chosen from a uniform distribution on $[0,1]$). Let $V_i$ be the solutions generated from running $\mathcal{A}$ for $k=n-i+1$ for $i=1, \ldots, n$. The cost of these solutions is determined by assigning each point to the nearest cluster center and is scaled so that the minimum cost for $k<n$ is at least one. We then order these solutions according to their cost into buckets of the form $[0,0], (\beta_0, \beta_0 \beta], (\beta_0 \beta, \beta_0 \beta^2], \ldots $.

From each non-empty bucket, pick the solution with the smallest number of cluster centers, and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r$. We then inductively build our nested solutions. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we define $S_i$ by iteratively adding the closest point $j \in S_{i-1}$ to $l$ for all $l \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

We then define an assignment $a_i$. For each $j \in X$ such that $a_{i-1}(j) \in S_{i}$, we set $a_i(j) = a_{i-1}(j)$. Otherwise, we set $a_i(j)$ to be the facility $i \in S_i$ such that assigning all points in $a_{i-1}^{-1}(j)$ (i.e. all points in $j$'s cluster in $S_{i-1}$) would have minimum cost. Using this assignment, $(a_i, S_i)$ is nested in $(a_{i-1}, S_{i-1})$.

We continue this nesting until $|S_i| = 1$, reaching the top level of the hierarchy. After forming this hierarchical clustering, for any $k \in 1, \ldots, n$ \cite{Lin} guarantee that the cost of $S_i$ is at most $20.71$ times the cost of $V_k$ where $|S_{i-1}| \geq n-k+1 \geq |S_{i}|$ (in the randomized case the expected cost of $S_i$ is at most $10.03$ times the cost of $V_k$).

\subsection{Flat Clustering Algorithms}
This nesting framework allows us to choose the flat clustering algorithm $\mathcal{A}$. As mentioned above, PAM and $k$-medoids are two potential algorithms. PAM has the added benefit of being a 5-approximation algorithm for the flat $k$-median problem which yields to an overall approximation ratio of $20.71 \cdot 5 = 103.55$ for the hierarchical $k$-median problem ($10.03 \cdot 5 = 50.15$ randomized). We also considered the size of the local swaps in PAM to improve the approximation ratio but found that the approximation ratio of constructed solutions for given values of $k$ did not significantly improve while the computation time increased greatly.

We also consider two linear programming based algorithms, one of which has an approximation guarantee. We first present the linear programming formulation for the $k$-median problem. Suppose we have a set $X$ and we let $s_i \in \{0,1\}$ be the variable representing whether or not we want to make $i$ a center for all $i \in X$. We also create a variable $a_{i,j} \in \{0, 1\}$ which represents whether or not we assign $i \in X$ to the cluster with center $j \in X$. Given this set and assignment, the $k$-median objective function is 
\begin{equation}
\sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j), 
\end{equation}
which we want to minimize. However, we are restricted by the fact that we can choose at most $k$ centers, every point must be assigned to a single cluster center, and that we can only assign a point to a center that is actually in $S$. This is represented by the three equations:
\begin{eqnarray}
\sum_{i \in X} s_i   \leq k & ,\\
\sum_{j \in X} a_{i,j}  = 1 & \forall i \in X, \\
a_{i,j}   \leq s_i & \forall i, j \in X.
\end{eqnarray}
Therefore, solving the following integer program would directly find the optimal set of centers and assignment for the $k$-median problem.
\begin{eqnarray}
\text{Min } & \sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j) \\
\text{s.t. } & \sum_{i \in X} s_i   \leq k  \\
& \sum_{j \in X} a_{i,j}  = 1 \quad \forall i \in X \\
& a_{i,j}   \leq s_i \quad \forall i, j \in X \\
& s_i \in \{0,1\} \quad \forall i \in X \\
& a_{i,j} \in \{0,1\} \quad \forall i, j \in X
\end{eqnarray}
Relaxing the constraints so that $s_i \in [0,1]$ for all $i \in X$ and $a_{i,j} \in [0,1]$ for all $i,j \in X$ yields the linear programming (LP) relaxation for the $k$-median problem, which we can solve in polynomial time.

Interestingly, the solution to the linear programming is often integer valued [NEED CITATION]. Based on this fact, we proposed a flat clustering algorithm that solves the LP relaxation and constructs $S$ by taking the points in $X$ with the $k$ highest $s_i$ values. We refer to this algorithm as LP-greedy from now on.
Charikar and Li \cite{charikar} introduce a 3.25-approximation algorithm that uses this LP solution to smartly construct a set of centers. We refer to this algorithm as LP-Charikar. We choose to implement this LP algorithm rather than the LP-based approximation algorithm with best approximation ratio because the algorithm is simpler to implement (and will most likely have a lower running time).

The algorithm works by first choosing a subset of data points $C \subseteq X$ to represent the demand. For each data point $i$, let $d_{av}(i) = \sum_{j \in X : a_{i,j} > 0} s_i \cdot d(i,j)$. We start with $C = \emptyset$. At each step, we choose a point $j \in X$ to add to $C$, and all points $j'$ such that $d(j,j') \leq 4 d_{av}(j') $ are removed from $X$. This is continued until $X$ is empty.

$C$ now represents data points that are relatively far apart. These points then choose a bundle of points $U_j$ that represent potential centers they would like to be assigned to. In particular, for each point $j \in C$, we look at all points $i \in X$ such that $0 < d(i,j) < 1.5 R_j$, where $R_j$ is half the distance from $j$ to it's nearest point in $C$, and bundle these points into $U_j$ (if a point is in multiple bundles you choose one arbitrarily). Now these bundles of points will have a sum of $s_i$ values at least one half. Lastly, these bundles are matched together greedily by distance. For each $U_j$, we let $\mathrm{vol}(U_j) = \sum_{i \in U_j} s_i$.

Now that the algorithm has grouped together points so the volume is at least 1, the algorithm starts to choosing points from these various groups through sampling. Here, opening $U_j$ refers to selecting a point $i \in U_j$ with probabilities relative to the $s_i$ values and adding that point to $S$. 
\begin{enumerate}[\IEEEsetlabelwidth{3)}]
\item For each matched $U_j$ and $U_{j'}$, the algorithm opens $U_j$ with probability $1-\mathrm{vol}(U_{j'})$, opens $U_{j'}$ with probability $1- \mathrm{vol}(U_j)$, and opens both $U_{j}$ and $U_{j'}$ with probability $\mathrm{vol}(U_j) + \mathrm{vol}(U_{j'}) -1$. 
\item For any unmatched $U_j$, it is opened with probability $\mathrm{vol}(U_j)$.
\item For any $i$ not in any $U_j$ such that $s_i > 0$, add $i$ to $S$ with probability $s_i$. 
\end{enumerate}
This sampling process returns a set of centers of size $k$  in expectation. For our purposes, we sample until we obtain a solution with exactly $k$ centers.

\section{Implementation and Results}
We implemented the nesting hierarchical framework above in Python 2.7, offering options for both the randomized and deterministic versions. For the black-box $k$-median algorithm, we implemented $k$-medoids, PAM, LP-greedy, and LP-charikar in Python using Gurobi 6.0.3 to solve the linear programming instances. We compared the results to hierarchical agglomerative clustering with both average linkage and Ward's linkage (implemented in the scipy package in Python).

To avoid cycling, we limited the number of iterations in $k$-medoids to 1000. For PAM, we also needed to limit the number of swaps to achieve a reasonable runtime. Instead of searching over all potential swaps, we sample at most $n$ potential ones. We also ensure that some minimum level progress is achieved by ensuring the algorithm improves by at least a factor of 0.9 on each iteration. At the end of PAM, we run 10 iterations of $k$-medoids to ensure the algorithm finds a local optimum. These choices were made by looking at the relative performance and computation time on random Gaussian data not tested below. For PAM and $k$-medoids, when solving for multiple values of $k$, we use the solution with $k$ centers as a warm start for $k+l$ centers by choosing an additional random $l$ centers. For the two LP algorithms, Gurobi uses a warm start from solving the LP for $k+l$ to solve it for $k$ because we found this reverse direction to be faster in practice.

We also tested a more computationally friendly heuristic of the hierarchical framework which only buckets solutions for values of $k$ that are a power of two. This yields many fewer $k$-median instances to solve, reducing the run-time significantly. This was motivated by the fact that the difference between various $k$-median solutions will be larger when $k$ is small so we expect our sampling to still represent much of the underlying structure.

For each data set and each clustering method, we report the approximation ratio, computation time, and FScore measure.  FScore is a measure that looks at the clusters represented on each level and how they represent the true clustering in terms of precision and recall. See \cite{Larsen} for more information. 
[ADD DEFINITION OF FSCORE AND MORE DETAIL]

[MAKE CODE AVAILABLE PUBLICLY]

Tests were conducted using Python 2.7.9 on a 64-bit Windows platform, on a 4 core 3.4 Ghz Intel i7 processor with 16 GB physical memory. Results are given in Tables 2-4  below with some results averaged due to space constraints. The results reported are average values across 10 trials.

The data sets tested against were:
\begin{LaTeXdescription}
\item[Gaussian] 27 synthetic $k$-center Gaussian data sets, generated in Python. In particular, Gauss\textunderscore d\textunderscore k\textunderscore e was generated by first choosing $k$ random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \cdot \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] \cite{Beasley} From a collection of test data sets created by Beasley with 40 test sets of size $100, 200, \ldots, 900$ and varying values of $k$. 
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{LaTeXdescription}

\begin{table}[!t]
\caption{Some specifics about our datasets}
\label{table_specifics}
\centering
\begin{tabular}{ | l | l | l | l | l | }
\hline
 Dataset Name & Opt Obj & $k$ & $n$ & dim \\ \hline
Soybean & 114 & 4 & 47 & 35 \\ 
Iris & 98 & 3 & 150 & 4 \\ 
Gauss\_5\_5\_5 & 195.4 & 5 & 388 & 5 \\ 
Gauss\_5\_10\_5 & 270 & 10 & 544 & 5 \\ 
Gaussian avg & 241 & 22 & 507 & 5 \\ 
pmed5 & 1355 & 33 & 100 & \  \\ 
pmed20 & 1789 & 133 & 400 & \  \\ 
% pmed40 & 5128 & 90 & 900 & \  \\ 
pmedian avg & 5535 & 46 & 460 & \  \\ \hline
 \end{tabular}
\end{table}

\begin{table*}[!t]
\label{fig:hier_results_comp}
\caption{Average Computation Time (sec), FScore measure, and Approximation Coeffiecients for Hierarchical Clustering Algorithms. K-med and PAM represent the black-box $k$-median solvers used and rand and logn indicate whether the randomized and logarithmic setting were used, respectively.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & Time & Time & Time & Time & Time & Time \\ \hline
	Dataset & Ave Linkage & Ward Linkage & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 0.0004 & 0.0002 & 0.286 & 0 & 0.294 & 0.167 \\ 
	Iris & 0.0021 & 0.0011 & 3.161 & 0.226 & 3.794 & 5.122 \\ 
	Gauss\_5\_5\_5 & 0.0097 & 0.0109 & 106.619 & 2.735 &120.893 & 147.137 \\
	Gauss\_5\_10\_5 & 0.0237 & 0.0256 & 383.854 & 10.259 & 383.298 & 494.554 \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.0229 & 0.0239 & 354.459 & 7.492 & 369.967 & 423.293 \\ 
	pmed5 &  0.0004 & N/A & 1.184 & 0.102 & 1.288 & 1.673   \\ 
	pmed25 & 0.0174 & N/A & 279.334 & 5.011 & 306.684 & 312.503   \\ 
	pmedian avg & 0.0067 & N/A &  87.355 & 1.957 & 94.548 & 101.576   \\ \hline
\end{tabular}
\end{table*}

\begin{table*}[!t]
\label{fig:hier_results_FScore}
\caption{Average FScore measure for Hierarchical Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & F-Score & F-Score & F-Score & F-Score & F-Score & F-Score \\ \hline
	Dataset & Ave Linkage & Ward Linkage & K-med & K-med logn rand & K-med rand & PAM logn rand \\ \hline
	Soybean & 0.615 & 0.554 & 0.772 & 0.643 & 0.742 & 0.633 \\ 
	Iris & 0.5 & 0.5 & 0.888 & 0.767 & 0.762 & 0.732 \\ 
	Gauss\_5\_5\_5 & 0.519 & 0.520 & 0.591 & 0.816 & 0.761 & 0.785 \\
	Gauss\_5\_10\_5 & 0.298 & 0.281 & 0.589 & 0.584 & 0.576 & 0.446 \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.393 & 0.378 & 0.649 & 0.616 & 0.624 & 0.624 \\ \hline
\end{tabular}
\end{table*}

\begin{table*}[!t]
\label{fig:hier_results_approx}
\caption{Average Approximation Coeffiecients for Hierarchical Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | }
\hline
	   & Approx & Approx & Approx & Approx \\ \hline
	Dataset & K-med & K-med logn rand & K-med rand & PAM logn rand \\ \hline
	Soybean & 1.585 & 2.334 & 1.711 & 2.001 \\ 
	Iris & 2560631.459 & 2550279.566 & 2787312.009 & 1623132.096 \\ 
	Gauss\_5\_5\_5 & 6.459 & 13.394 & 8.624 & 7.791 \\
	Gauss\_5\_10\_5 &  3.420 & 3.756 & 4.196 & 2.309  \\
	%Gauss\_5\_50\_5 &  & & &  \\
	Gaussian Average & 5.182 & 7.262  & 5.537 & 4.739 \\ 
	pmed5 & 11.367 & 51.141 & 16.261 & 11.950 \\ 
	pmed25 & 6.641 & 13.372 & 5.378 & 13.200 \\ 
	pmedian avg & 10.330 & 19.822 & 10.524 & 8.599 \\ \hline
\end{tabular}
\end{table*}

In general, the hierarchical framework performed significantly better than both methods of agglomerative clustering in terms of FScore measure, even for our logarithmic version. The FScore measures were very low for the agglomerative clustering methods, sometimes around 0.25, but were much closer to the AMI measures we saw in flat clustering for our implemented algorithms.

For $k$-medoids in the non-logarithmic version and PAM in the logarithmic version, the hierarchical framework increased the computation time by three orders of magnitude, and increased by two orders of magnitude for $k$-medoids in the logarithmic version. Interestingly, the approximation factors did not change significantly between the randomized and deterministic versions for $k$-medoids. The approximation ratios were strongest as $n$ increased or as the dimension increased. In particular, for the Gaussian data in 7 dimensions, the approximation ratio was often around 3.  The high approximation ratio for the iris data set was due to the fact that the optimal solution for $k = n-1, n-2$ was very close to zero, and our heuristic $k$-median algorithm only sometimes picked up on these solutions. The approximation factor at lower levels was more consistent with the other results.

\section{Conclusion and Future Work}
We implemented various simple approximation algorithms for partitional clustering, and a framework for hierarchical clustering.  We compared these algorithms, along with heuristic modifications, against the most common clustering algorithms used in practice.

For partitional clustering, we found that the approximation algorithms in the literature are in general slower and cannot improve much on current heuristics used despite their approximation guarantees. This was not true in the hierarchical clustering case where we saw significant improvement in FScore measures for our implemented methods and overall good approximation ratios where none were had before. In particular, our logarithmic nesting algorithm seems to provide the best balance between performance and computational efficiency.

In the future, we hope to explore how linear programming based methods perform in both the flat and hierarchical context as well and to implement our own agglomerative clustering to help make smarter local merges in the nesting framework.

>>>>

\subsection{Hierarchical Algorithm}

Recall the hierarchical $k$-median problem: given a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$, we represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$.
 A solution to the hierarchical $k$-median problem is a chain of nested solutions $(S_1, a_1) \subseteq (S_2, a_2) \subseteq \ldots \subseteq (S_l, a_l)$. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the hierarchy.  
Consider any $k$-median algorithm $\mathcal{A}$. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce an augmentation step that nests these clusterings into the hierarchical framework and only increase costs by a bounded factor.

First, set $\beta = 3+\sqrt{3}$, and $\beta_0 = 1$ (for the randomized algorithm we switch $\beta = 6.355$ and set $\beta_0 = \beta^X$, where $X$ is chosen from a uniform distribution on $[0,1]$). Let $V_i$ be the solutions generated from running $\mathcal{A}$ for $k=n-i+1$ for $i=1, \ldots, n$. The cost of these solutions is determined by assigning each point to the nearest cluster center and is scaled so that the minimum cost for $k<n$ is at least one. We then bucket these solutions by cost into buckets of the form $[0,0], (\beta_0, \beta_0 \beta], (\beta_0 \beta, \beta_0 \beta^2], \ldots $.

From each non-empty bucket, pick the solution with the smallest number of cluster centers, and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r$. We then inductively build our nested solutions. Set $S_1 = \overline{V}_1$ and $a_1$ to be the assignment that assigns each point to the closest point in $S_1$. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we construct $S_i$ by  adding the closest point $j \in S_{i-1}$ to $l$ for each $l \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

\begin{figure*}[!t]
\centering
\subfloat[Nested solution $(S_{i-1}, a_{i-1})$ ]{\includegraphics[width =1.75in]{Si-1}
\label{fig_Si-1}} \hspace{15mm}
\subfloat[Next set $\bar{V}_i$]{\includegraphics[width=1.75in]{Vi}
\label{fig_Vi}} \\
\subfloat[Next nested centers $S_i$]{\includegraphics[width=1.75in]{Si}
\label{fig_Si}}  \hspace{15mm}
\subfloat[Next nested solution $(S_i, a_i)$]{\includegraphics[width=1.75in]{Siai}
\label{fig_Siai}}
\caption{An example of nesting $\bar{V}_i$ with $(S_{i-1}, a_{i-1})$. Dark  nodes represent chosen centers and edges represent assignments to that center's cluster. First, node 2 chooses node from 1 from $S_{i-1}$ as its closest center and node 5 chooses node 4. Then, each cluster in $(S_{i-1}, a_{i-1})$ is assigned to the cheapest (lowest sum of distances) center.}
\label{fig_sim}
\end{figure*}

We then need to define an assignment $a_i$. For each $j \in X$ such that $a_{i-1}(j) \in S_{i}$, we set $a_i(j) = a_{i-1}(j)$. Otherwise, we set $a_i(j)$ to be the facility $i \in S_i$ such that assigning all points in $a_{i-1}^{-1}(j)$ (i.e. all points in $j$'s cluster in $S_{i-1}$) would have minimum cost. Using this assignment, $(a_i, S_i)$ is nested in $(a_{i-1}, S_{i-1})$. An example of this nesting is given in Fig.~\ref{fig_Si-1}-\ref{fig_Siai}.

We continue this nesting until $|S_i| = 1$, reaching the top level of the hierarchy where all data points are in the same cluster. After forming this hierarchical clustering, for any $k \in 1, \ldots, n$ \cite{Lin} guarantees that the cost of $S_i$ is at most $20.71$ times the cost of $V_k$ where $|S_{i-1}| \geq |V_k| \geq |S_{i}|$ (in the randomized case the expected cost of $S_i$ is at most $10.03$ times the cost of $V_k$).

\subsection{Partitional Clustering Algorithms}
This nesting framework allows us to choose the partitional clustering algorithm $\mathcal{A}$. As mentioned above, PAM and $k$-medoids are two potential algorithms. PAM has the added benefit of being a 5-approximation algorithm for the partitional $k$-median problem which yields to an overall approximation ratio of $20.71 \cdot 5 = 103.55$ for the hierarchical $k$-median problem ($10.03 \cdot 5 = 50.15$ randomized). We also considered the size of the local swaps in PAM to improve the approximation ratio but found that the approximation ratio of constructed solutions for given values of $k$ did not significantly improve while the computation time increased greatly.

We also consider two linear programming based algorithms, one of which has an approximation guarantee. We first present the linear programming formulation for the $k$-median problem. Suppose we have a set $X$ and we let $s_i \in \{0,1\}$ be the variable representing whether or not we want to make $i$ a center for all $i \in X$. We also create a variable $a_{i,j} \in \{0, 1\}$ which represents whether or not we assign $i \in X$ to the cluster with center $j \in X$. Given this set and assignment, the $k$-median objective function is 
\begin{equation}
\sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j), 
\end{equation}
which we want to minimize. However, we are restricted by the fact that we can choose at most $k$ centers, every point must be assigned to a single cluster center, and that we can only assign a point to a center that is actually in $S$. This is represented by the three equations:
\begin{eqnarray}
\sum_{i \in X} s_i   \leq k & ,\\
\sum_{j \in X} a_{i,j}  = 1 & \forall i \in X, \\
a_{i,j}   \leq s_j & \forall i, j \in X.
\end{eqnarray}
Therefore, solving the following integer program would directly find the optimal set of centers and assignment for the $k$-median problem.
\begin{eqnarray}
\text{Min } & \sum_{i \in X} \sum_{j \in X} a_{i,j} d(i,j) \\
\text{s.t. } & \sum_{i \in X} s_i   \leq k  \\
& \sum_{j \in X} a_{i,j}  = 1 \quad \forall i \in X \\
& a_{i,j}   \leq s_j \quad \forall i, j \in X \\
& s_i \in \{0,1\} \quad \forall i \in X \\
& a_{i,j} \in \{0,1\} \quad \forall i, j \in X
\end{eqnarray}
Relaxing the constraints so that $s_i \in [0,1]$ for all $i \in X$ and $a_{i,j} \in [0,1]$ for all $i,j \in X$ yields the linear programming (LP) relaxation for the $k$-median problem, which we can solve in polynomial time.

Interestingly, the solution to the linear programming is often integer valued \cite{Nagarajan}. Based on this fact, we propose two partitional clustering algorithms: one that solves the LP relaxation and constructs $S$ by taking the points in $X$ with the $k$ highest $s_i$ values and one that solves the LP relaxation and constructs $S$ by choosing $i$ to be in $S$ with probability $s_i$, sampling for each $i \in X$ until $k$ centers are chosen. We refer to these algorithms as LP-greedy and LP-rand, respectively.
Charikar and Li \cite{Charikar} introduce a 3.25-approximation algorithm that uses this LP solution to smartly construct a set of centers. We refer to this algorithm as LP-Charikar. We choose to implement this LP algorithm rather than the LP-based approximation algorithm with best approximation ratio because the algorithm is simpler to implement (and we expect it to have a lower running time).

The algorithm works by first choosing a subset of data points $C \subseteq X$ to represent the demand. For each data point $i$, let $d_{av}(i) = \sum_{j \in X : a_{i,j} > 0} s_j \cdot d(i,j)$. We start with $C = \emptyset$. At each step, we choose a point $j \in X$ to add to $C$, and all points $j'$ such that $d(j,j') \leq 4 d_{av}(j') $ are removed from $X$. This is continued until $X$ is empty.

$C$ now represents data points that are relatively far apart. These points then choose a bundle of points $U_j$ that represent potential centers they would like to be assigned to. In particular, for each point $j \in C$, we look at all points $i \in X$ such that $0 < d(i,j) < 1.5 R_j$, where $R_j$ is half the distance from $j$ to its nearest point in $C$, and bundle these points into $U_j$ (if a point is in multiple bundles, choose one arbitrarily). Now these bundles of points have a sum of $s_i$ values of at least one half. For each $U_j$, we define the volume of $U_j$ to be $\mathrm{vol}(U_j) = \sum_{i \in U_j} s_i$. Lastly, these bundles are matched together greedily by distance.

Now that the algorithm has grouped and matched together points so the volume is at least 1, the algorithm starts choosing points from these various groups through sampling. Here, opening $U_j$ refers to selecting a point $i \in U_j$ with probabilities proportional to the $s_i$ values and adding that point to $S$. 
\begin{enumerate}[\IEEEsetlabelwidth{3)}]
\item For each matched $U_j$ and $U_{j'}$, the algorithm opens $U_j$ with probability $1-\mathrm{vol}(U_{j'})$, opens $U_{j'}$ with probability $1- \mathrm{vol}(U_j)$, and opens both $U_{j}$ and $U_{j'}$ with probability $\mathrm{vol}(U_j) + \mathrm{vol}(U_{j'}) -1$. 
\item For any unmatched $U_j$, it is opened with probability $\mathrm{vol}(U_j)$.
\item For any $i$ not in any $U_j$ such that $s_i > 0$, add $i$ to $S$ with probability $s_i$. 
\end{enumerate}
This sampling process returns a set of centers of size $k$  in expectation and produces a set $S$ such that $\mathbb{E}[f(S)] \leq 3.25 \cdot f(O)$ for all $O \subseteq X$ such that $|O| = k$. For our purposes, we sample until we obtain a solution with exactly $k$ centers.

\section{Implementation and Results}

\subsection{Implementation Specifics}

All tests were conducted using Python 2.7.9 on a 64-bit Windows platform with a 4 core 3.4 Ghz Intel i7 processor and 16 GB physical memory. All code is available at \url{https://github.com/apaul29/Hierarchical-Code}.
We chose to test our methods on a variety of data:
\begin{LaTeXdescription}
\item[Gaussian] 27 synthetic $k$-center Gaussian data sets, generated in Python. In particular, Gauss\textunderscore d\textunderscore k\textunderscore e was generated by first choosing $k$ random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \cdot \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] \cite{Beasley} From a collection of classic test data sets for approximation algorithms created by Beasley with 40 test sets of size $100, 200, \ldots, 900$ and varying values of $k$. This data is unlabeled.
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{LaTeXdescription}

\begin{table}[!t]
\caption{Some specifics about our datasets}
\label{table_specifics}
\centering
\begin{tabular}{ | l | l | l | l | l | }
\hline
 Dataset Name & Opt Obj & $k$ & $n$ & dim \\ \hline
Soybean & 114 & 4 & 47 & 35 \\ 
Iris & 98 & 3 & 150 & 4 \\ 
Gauss\_5\_5\_5 & 195.4 & 5 & 388 & 5 \\ 
Gauss\_5\_10\_5 & 270 & 10 & 544 & 5 \\ 
Gaussian avg & 241 & 22 & 507 & 5 \\ 
pmed5 & 1355 & 33 & 100 & \  \\ 
pmed20 & 1789 & 133 & 400 & \  \\ 
% pmed40 & 5128 & 90 & 900 & \  \\ 
$p$-median avg & 5535 & 46 & 460 & \  \\ \hline
 \end{tabular}
\end{table}

\subsection{Partitional Algorithms}
 For the black-box $k$-median algorithm, we implemented $k$-medoids, PAM, LP-greedy, LP-rand, and LP-Charikar in Python using Gurobi 6.0.3 to solve the linear programming instances. 
To avoid cycling, we limited the number of iterations in $k$-medoids to 1000. For PAM, we also needed to limit the number of swaps to achieve a reasonable runtime. Instead of searching over all potential swaps, we sample at most $n$ potential ones. We also ensure that some minimum progress is achieved by ensuring the algorithm decreases in objective by at least a factor of 0.9 in each iteration. At the end of PAM, we run 10 iterations of $k$-medoids to ensure the algorithm finds a local optimum. These choices were made by looking at the relative performance and computation time on random Gaussian data not in the data sets above.

Before implementing these data sets in the hiearchical framework, we tested them as partitional algorithms on the data above which had given values of $k$. For comparison, we also included the results from running $k$-means (implemented in the scikit-learn package in Python) and the extension of PAM called 2-swaps that also looks at local swaps of size 2 (which improves the $k$-median approximation ratio to 4 from 5).

For each data set and each partitional clustering method, we recorded the approximation ratio (for the flat $k$-median objective function), computation time, and, where applicable, adjusted mutual information (AMI).  AMI is a measure which scores accuracy of predicted clusters compared to a true labeling by measuring the agreement of two assignments. The measure is normalized to reflect chance, so that a perfect labeling receives a score of 1 and an independent random labeling receives a score of 0 in expectation.  See \cite{Vinh} for further information.  Results are given in Tables~\ref{fig:flat_hier_comp}-\ref{fig:flat_hier_approx}  below with some results averaged due to space constraints. The results reported are average values across 10 trials.

\begin{table*}[!t]
\label{fig:flat_results_comp}
\caption{Average Computation Time (sec) for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l|}
\hline
Dataset	&	k-means	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	0.0151	&	0.0020	&	0.0159	&	0.0493	&	0.1180	&		&		\\
Iris	&	0.0129	&	0.0030	&	0.1018	&	0.2294	&	1.4644	&		&		\\
Gauss\_5\_5\_5	&	0.0249	&	0.0318	&	1.5058	&	2.3824	&	18.1701	&		&		\\
Gauss\_5\_10\_5	&	0.0540	&	0.1085	&	5.4071	&	9.4277	&	47.7316	&		&		\\
Gauss\_5\_50\_5	&	0.1371	&	0.3019	&	21.4457	&	39.6144	&	17.5115	&		&		\\
Gaussian avg	&	0.0730	&	0.1817	&	13.0214	&	18.6957	&	37.2315	&		&		\\
pmed5	&	N/A	&	0.0381	&	0.6223	&	0.7479	&	0.5357	&		&		\\
pmed20	&	N/A	&	1.7947	&	44.9791	&	68.2306	&	18.2055	&		&		\\
pmed35	&	N/A	&	0.1133	&	5.8814	&	9.2881	&	117.4038	&		&		\\
$p$-median avg	&	N/A	&	0.8072	&	40.3474	&	56.7595	&	46.1541	&		&		\\ \hline
\end{tabular}
\end{table*}

\begin{table*}[!t]
\label{fig:flat_results_AMI}
\caption{Average AMI measure for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l | l | l| }
\hline
Dataset	&	k-means	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	0.686	&	0.550	&	0.732	&	0.708	&	0.721	&		&		\\
Iris	&	0.748	&	0.690	&	0.760	&	0.757	&	0.647	&		&		\\
Gauss\_5\_5\_5	&	0.741	&	0.687	&	0.749	&	0.744	&	0.754	&		&		\\
Gauss\_5\_10\_5	&	0.783	&	0.677	&	0.725	&	0.737	&	0.729	&		&		\\
Gauss\_5\_50\_5	&	0.646	&	0.621	&	0.619	&	0.631	&	0.631	&		&		\\
Gaussian avg	&	0.777	&	9.112	&	0.761	&	0.761	&	0.767	&		&		\\ \hline
\end{tabular}
\end{table*}

\begin{table*}[!t]
\label{fig:flat_results_approx}
\caption{Average Approximation Coeffiecients for Partitional Clustering Algorithms.}
\centering
\begin{tabular}{ | l | l | l | l | l | l| l|}
\hline
Dataset	&	k-medoids	&	PAM	&	2-swaps	&	LP-Charikar	&	LP-greedy	&	LP-rand	\\ \hline
Soybean	&	1.167	&	1.012	&	1.012	&	1.000	&		&		\\
Iris	&	1.081	&	1.004	&	1.004	&	0.873	&		&		\\
Gauss\_5\_5\_5	&	1.070	&	1.002	&	1.005	&	1.002	&		&		\\
Gauss\_5\_10\_5	&	1.080	&	1.011	&	1.009	&	1.020	&		&		\\
Gauss\_5\_50\_5	&	1.089	&	1.020	&	1.014	&	1.001	&		&		\\
Gaussian avg	&	1.064	&	1.009	&	1.008	&	1.004	&		&		\\
pmed5	&	1.373	&	1.026	&	1.033	&	1.000	&		&		\\
pmed20	&	1.501	&	1.032	&	1.027	&	1.000	&		&		\\
pmed35	&	1.083	&	1.005	&	1.007	&	1.027	&		&		\\
$p$-median avg	&	1.187	&	1.017	&	1.019	&	1.000	&		&		\\ \hline
\end{tabular}
\end{table*}

[ADD FLAT RESULT ANAYLIS]

It is interesting to note that, for the Gaussian data, an increase in dimension, reduction in $k$, or reduction in the percentage of noisy data made it easier for the algorithms to classify the data set, which makes sense based on our intuition of clustering. For the OR library instances, we saw that the approximation ratio tended to increase as $k$ grew larger but not as $n$ grew.
